---
title: 'Reproducible Research: Peer Assessment 2'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

# What weather events influence to economics and public health 
The aim of this report is to explor U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and answer to the following questions
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?
During data analisys we identified weather events with the greatest consequences w.r.t to public health and economy based on number of victims and economic loss.

## Data Processing
U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database
which is explored in this report is available by the link [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb].
This database tracks characteristics of major storms and weather events in
the United States, including when and where they occur, as well as estimates of
any fatalities, injuries, and property damage.


### Reading the data
Storm database is a CSV file compressed by bzip2 algorithm.
We read it as is and explore its content.
```{r read_data, cache=TRUE}
storm.data <- read.csv("stormdata.csv.bz2", na.strings="")
```

Show first few rows of the data file.
```{r show_rows, cache=TRUE}
dim(storm.data)
head(storm.data)
```

### Preparing data for analisys
From the storm databes we select only those columns which help to determine
economic and health consequeties of weather events. Examining the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
and the actual data (see data content above) it's seen that *FATALITIES* and
 *INJURIES* columns correspond to public health, columns *PROPDMG*, *PROPDMGEXP*,
 *CROPDMG* and *CROPDMGEXP* are related to econmic consequences of the weather events.

```{r cache=TRUE}
library(dplyr)
storm.cons <- tbl_df(storm.data) %>% select(STATE__, BGN_DATE, STATE, EVTYPE, FATALITIES,INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

head(storm.cons)
```

### Converting units
Property and crop damages are presented in dollars as numeric value and its
 magnitude ("1.55 B" - 1.55 billion dollars)
Available magnitude levels are *K* for thousands, *M* for millions, *B* for billions.
As we see in actual data has more than these three levels:
```{r}
unique(storm.cons$PROPDMGEXP)
unique(storm.cons$CROPDMGEXP)
```

From the Storm Data Documentation meaning of the rest of the levels is unclear.
So all unknown magnitued levels will be ignored. Only numeric part will be considered.

```{r convert_units, cache=TRUE}
convert.magnitude <- function(value) {
    if (is.na(value)) {
        1
    } else if (value == "K" | value == "k") {
        1000
    } else if (value == "M" | value == "m") {
        1000000
    } else if (value == "B" | value == "b") {
        1000000000
    } else {
        1
    }
}

storm.cons$PROPDMGEXP <- sapply(storm.cons$PROPDMGEXP, convert.magnitude)
storm.cons$CROPDMGEXP <- sapply(storm.cons$CROPDMGEXP, convert.magnitude)
storm.cons <- storm.cons %>% mutate(PROPDMG=PROPDMG*PROPDMGEXP,
CROPDMG=CROPDMG*CROPDMGEXP) %>% select(-PROPDMGEXP, -CROPDMGEXP)
```

### Imputing missing values
There is no missing values in columns that are needed for analisys.
```{r imputing_missing_values}
sum(is.na(storm.cons$PROPDMG))
sum(is.na(storm.cons$CROPDMG))
sum(is.na(storm.cons$FATALITIES))
sum(is.na(storm.cons$INJURIES))
```

## Results

### Most harmful weather events with respect to population health acrross the United States
Consider number of victims as a weather event effect on population health.
Number of victims is a sum of fatalities and injures for each event type.
```{r most_harmful_weather_events, cache=TRUE}
ph <- storm.cons %>%
    mutate(VICTIMS=FATALITIES+INJURIES) %>%
    group_by(EVTYPE) %>%
    summarise(VICTIMS=sum(VICTIMS)) %>%
    arrange(desc(VICTIMS))
```

Top 10 most harmfull weather events with respect to population health
```{r}
head(ph,n=10)
```
As seen from above table the weather event most harmfult w.r.t public health is *tornado*

Plot below shows victims of tornado through the whole history of observations (1950-2011).
Red line on the plot shows the median of victims.
```{r victims_of_tornado, cache=TRUE}
library(lubridate)
tornado <- storm.cons %>%
    mutate(YEAR=year(mdy_hms(BGN_DATE))) %>%
    filter(EVTYPE=="TORNADO") %>%
    mutate(VICTIMS= FATALITIES + INJURIES) %>%
    select(YEAR, VICTIMS) %>%
    group_by(YEAR) %>%
    summarise(VICTIMS=sum(VICTIMS))

plot(tornado, main="Victims of tornado across the United States")
abline(h=median(tornado$VICTIMS), col="red")
```

### Weather events with the greatest economic consequneses across the United States
Economic consequenses caused by weather events are of two types property damages and crop demages.
Sum of these two factors is a measure of weather event's efffect to economics.

```{r economic_loss, cache=TRUE}
economic.loss <- storm.cons %>%
    mutate(ECONOMIC.LOSS=(PROPDMG+CROPDMG)/1000000) %>%
    group_by(EVTYPE) %>%
    summarise(ECONOMIC.LOSS=sum(ECONOMIC.LOSS)) %>%
    arrange(desc(ECONOMIC.LOSS))
```

Below is the top 10 weather event type with the greatest economic consequenses. Values are in millions of dollars.
```{r}
head(economic.loss, n=10)
```
Let's see how weather events affects the economic of the United States fot the past ten years

```{r cache=TRUE}
past10 <- storm.cons %>%
    mutate(YEAR=year(mdy_hms(BGN_DATE))) %>%
    filter(YEAR > 1990) %>%
    mutate(ECONOMIC.LOSS=(PROPDMG+CROPDMG)/1000000) %>%
    select(YEAR, ECONOMIC.LOSS) %>%
    group_by(YEAR) %>%
    summarise(ECONOMIC.LOSS=sum(ECONOMIC.LOSS))

plot(past10, main="Economic loss for various weather events for past 10 years", ylab="Economic loss (millions of dollars)")
```
